
* Fraction of ERCC-originating reads: As mentioned above high fractions of ERCC-originating reads are also indicative of low cell quality. We already saw that 
one batch  (NA19098.r2) appeared particurarly problematic in this respect and we shall therefore remove it.
```{r filter_high_ERCC}
high_ERCC=which(anno[,4]=="NA19098.r2")
```


We will try PCA without any filter for dispersion but feeding in two different transformations of the data; A simple log transformation of the count matrix and the cell to cell correlation matrix.
First let's perform PCA on the two versions of the data and look at the proportion of explained variance by the first 50 components:

```{r PCA_no_selection}
endog=setdiff(rownames(clean_norm_umi_counts),ERCC) #Use only endogenous genes (remove ERCCs)
LogTransf_counts=log2(clean_norm_umi_counts[endog,]+1)
PCA_log=prcomp(x=t(LogTransf_counts),scale=T) #Note that for prcomp we need to pass a trnasposed version of the matrix where cells are the rows and genes the columns
PCA_cor=prcomp(x=cor(log2(clean_norm_umi_counts[endog,]+1)),scale=T) 
par(mfrow=c(1,2))
plot(summary(PCA_log)$importance[3,1:50],type="l",xlab="#PC (LogTrans. Data)",ylab="Cum. Proportion of Explained Variance")
plot(summary(PCA_cor)$importance[3,1:50],type="l",xlab="#PC (Cell2Cell cor.)",ylab="Cum. Proportion of Explained Variance")
Log_NODG=log2(colSums(clean_norm_umi_counts>0)+1) # Calculate also the number of detected genes (to use later in plotting)
datt=data.frame(logNODG=Log_NODG,Batch=clean_anno[,4],PCA_log=PCA_log$x[,1:2],PCA_cor=PCA_cor$x[,1:2])  # to ease plotting combine results in a dataframe 
```

Now let's project...
First using the PCA on the Log transformed data:
  ```{r PCA_no_selection_plot1}
chart_logNODG=ggplot(datt,aes(x=PCA_log.PC1,y=PCA_log.PC2,color=logNODG))
chart_logNODG=chart_logNODG+scale_color_gradient(low="#FF5544",high="#4455FF")
chart_logNODG=chart_logNODG+geom_point(size=4,alpha=0.8)
chart_batch=ggplot(datt,aes(x=PCA_log.PC1,y=PCA_log.PC2,color=Batch))
chart_batch=chart_batch+geom_point(size=4,alpha=0.8)
grid.arrange(chart_logNODG, chart_batch, ncol=1)
```
And then using the PCA on the cell2cell correlation:
  ```{r PCA_no_selection_plot2}
chart_logNODG=ggplot(datt,aes(x=PCA_cor.PC1,y=PCA_cor.PC2,color=logNODG))
chart_logNODG=chart_logNODG+scale_color_gradient(low="#FF5544",high="#4455FF")
chart_logNODG=chart_logNODG+geom_point(size=4,alpha=0.8)
chart_batch=ggplot(datt,aes(x=PCA_cor.PC1,y=PCA_cor.PC2,color=Batch))
chart_batch=chart_batch+geom_point(size=4,alpha=0.8)
grid.arrange(chart_logNODG, chart_batch, ncol=1)
```


* PCA after selecting for overdispersed genes. 

We will now see what is the effect on selecting overdispersed genes prior to PCA. Let's select the top 25% overdispersed genes and perform PCA on this subset of our dataset:
```{r PCA_selection}
genes_keep <- select_variable_genes(clean_norm_umi_counts,0.25) #Select the top 25% of overdispersed genes
endog=setdiff(rownames(clean_norm_umi_counts[genes_keep,]),ERCC) #Use only endogenous genes (remove ERCCs)
LogTransf_counts=log2(clean_norm_umi_counts[endog,]+1)
PCA_log=prcomp(x=t(LogTransf_counts),scale=T) #Note that for prcomp we need to pass a trnasposed version of the matrix where cells are the rows and genes the columns
PCA_cor=prcomp(x=cor(log2(clean_norm_umi_counts[endog,]+1)),scale=T) 
par(mfrow=c(1,2))
plot(summary(PCA_log)$importance[3,1:50],type="l",xlab="#PC (LogTrans. Data)",ylab="Cum. Proportion of Explained Variance")
plot(summary(PCA_cor)$importance[3,1:50],type="l",xlab="#PC (Cell2Cell cor.)",ylab="Cum. Proportion of Explained Variance")
datt=data.frame(logNODG=Log_NODG,Batch=clean_anno[,4],PCA_log=PCA_log$x[,1:2],PCA_cor=PCA_cor$x[,1:2])  # to ease plotting combine results in a dataframe 
```


Let's project again...
First using the PCA on the Log transformed data:
  ```{r PCA_selection_plot1}
chart_logNODG=ggplot(datt,aes(x=PCA_log.PC1,y=PCA_log.PC2,color=logNODG))
chart_logNODG=chart_logNODG+scale_color_gradient(low="#FF5544",high="#4455FF")
chart_logNODG=chart_logNODG+geom_point(size=4,alpha=0.8)
chart_batch=ggplot(datt,aes(x=PCA_log.PC1,y=PCA_log.PC2,color=Batch))
chart_batch=chart_batch+geom_point(size=4,alpha=0.8)
grid.arrange(chart_logNODG, chart_batch, ncol=1)
```
And then using the PCA on the cell2cell correlation:
  ```{r PCA_selection_plot2}
chart_logNODG=ggplot(datt,aes(x=PCA_cor.PC1,y=PCA_cor.PC2,color=logNODG))
chart_logNODG=chart_logNODG+scale_color_gradient(low="#FF5544",high="#4455FF")
chart_logNODG=chart_logNODG+geom_point(size=4,alpha=0.8)
chart_batch=ggplot(datt,aes(x=PCA_cor.PC1,y=PCA_cor.PC2,color=Batch))
chart_batch=chart_batch+geom_point(size=4,alpha=0.8)
grid.arrange(chart_logNODG, chart_batch, ncol=1)
```
